From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-31 appendixB/appendixB.Rmd
WIP

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}.}

@@

>>>>

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}

@@

# System size expansion in higher-dimensional systems: environmental noise example

We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.  It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.  Following the expansion as before (see @vanKampen for details), the mean dynamics are given by the macroscopic equations:

\begin{align}
\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) \\
\end{align}

while the fluctuation dynamics are now given by three coupled differential equations:

\begin{align}
\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial \alpha_{1,0}}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \xi \eta \rangle + \alpha_{2,0} \label{phivar} \\
\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial \alpha_{1,0}}{\partial \phi} + \frac{\partial \beta_{1,0} }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_{1,0}}{\partial \psi} \langle \eta^2 \rangle + \beta_{2,0} \label{psivar}
\end{align}

Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_{1,0}}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_{2,0}}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.  A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.

Setting the remaining equations to zero to solve for the stationary state we can write down a general formula for the variation in $\phi$:

\begin{equation}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation}

From which we recover the environmental variation equation presented in the main text.

### 2017-10-31 paper.Rmd
WIP

What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model,
it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early quantiative ecological models first formulated at the population level
as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
any random component as mere noise [@Coulson2004]. Advances in both theoretical understanding and the advent of widely available and
efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
into quantiative models [@Black2012; @Ovaskainen2010].

@@

The van Kampen system size expansion provides a convenient way to explicitly approximate the "macroscopic" (i.e. population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].  In addition to illustrating how common stochastic models (in particular, Guassian noise) arise from lower-level discriptions, this result will also provide us a useful framework for interpeting the behavior of stochastic models more generally.  In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:

@@

Note that like the equation for the mean, we have a dynamic equation for the time-evolution of the variance in Eq \eqref{bd}.  We can solve these equations at steady-state by setting both equations zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.  We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.  For instance, in a logistic model with $b = rn$ and $d = rn^2/N$, we instead find the variance at steady state is $\sigma^2 = K$; significantly larger than in the Levins model.

@@

$$\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d$$

where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the autocorrelation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:

@@

Note that the contribution of environmental stochasticity is independent of system size and directly proprotional to the variance in the environment.  This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$.  The correlation time $\tau_c$, is really a shorthand for $\tfrac{1}{\partial_y g(y)}$.  Coulpling

<!--
[@Roughgarden; @Ellner]

[@Ovaskainen2010] Master equation, break-down of Fokker-Planck approximation under large noise, introduction of the WKB approximation
[@Bjornstad2001] _Science_ Review of noisy fluctuations

Colored noise: [@Lawton]
- [@Ripa1999]
- [@Marshall2015]
-->

### Other forms of stochasticity

@@

A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species. Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guarenteed to eventually go extinct under demographic noise.  While extinction is certainly a phenonomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself. Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.  In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].

@@

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.  This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.  The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:

@@

where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.  The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.  This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.  The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g. @Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).  This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.  (Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).  Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.  While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.  While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.

@@

Despite the nearly ubiquitous association of a ball-in-cup model with ecological discussions of resilience and stability, this potential well diagram (Figure 3C) is also a source of confusion on the topic.  As we have seen, the most salient feature of this diagram is the depth of each well: transistion probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.  However, it is important to remember that because the potential function is defined by the above integral, it in fact already combines two more elementary and mechanistically precise quantities from the growth equation: the slope of the growth rate (stability) and the distance to the tipping point (size of the basin of attraction).  Both a steeper negative slope or a longer distance to integrate over between stable and unstable points will create a deeper well.

Empirical evidence for stochastic switching in ecological timeseries is limited by the long time scale usually required. Because a single stochastic transisition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013]. However, these transistions are most likely to observed in systems with relatively shallow potential wells.  This is mostly likely to occur of some external change is slowly making the potential well shallower and thus transitions more likely -- a phenomenon that has become to be known as "flickering" and suggested as a potential early warning sign of an approaching transition [@Brock2010].  Examples of flickering between alternative stable states have been suggested in data ranging from geological timescales at the end of glaciation in the Pleistocene [@Taylor1993] to ecological scales over a few decades in lake ecosystems [@Wang2012].  Stochastic transitions between bistable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonally reproducing microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].

@@

Noise can provide novel information in other ways as well.  Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pusing a system away from an equilibrium to reveal it's behavior elsewhere in state space. Observing the size and autocorrelation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-disspation result we found for generic birth death models).  This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifucation.  This pattern, known as critical slowing down (in reference to the autocorrelation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has recieved considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].  Increases in two of these early warning indicators, variance and autocorrelation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.  This simulation is based on a mechanistic model developed by @Dai2012 that closely matches the population dynamics they observed in serial dilutions of budding yeast _Saccharomyces cervisiae_ on a sucrose medium. Because the relevant proteins are located in the extra-cellular matrix rather than inside the cells, budding yeast break down sucrose collaboratively, creating an Allee effect below which the population is too small to sustain itself.  Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing autocorrelation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.

As we saw in the equations from the system-size approximation, the variance and autocorrelation in question are technically properties of an ensemble, and should be computed over replicate simulations.  While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.  If the environmental change is sufficiently slow relative to the measurements, one can assume that the system is close to its stationary state over a given window in time, computing these averages over a rolling time window as we illustrate here, rather than across an ensemble of replicates.

@@

The past two decades have seen widespread recognition of the central role that stochasticity plays through the use of mechanistic models can have in creating and understanding ecological phenomena [@Coulson2004; @Black2012]. We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convienient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].  These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought determinstic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.  Yet too many of these advances have remained limited to the domain of theory; limited to deductive models which may seek to mimic observed patterns but are rarely used to confront empirical data directly.  There is good reason for this: statistical inference of even relatively simple models with multiple sources of noise will always be difficult given limited ecological data [@Knape2011].  However, many stochastic phenomena may suggest new ways to infer rather than merely obscure underlying processes.

The eruption of interest in early warning signals for critical transitions [@Scheffer2009] provides an excellent illustration of this potential.  A range of stochastic phenomena have been suggested and a tested for their potential to reveal subtle signs of ecosystem change prior to a dramatic shift, and prompted a rich dialog between further theory and new experiment. Changes in stochastic properties such as variance or autocorrelation provide evidence of resilience loss that would be invisible to a determinstic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transistion.  In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of emprical investigation and further theoretical work to test the potential of this approach.  The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transistion or fold bifurcation.

Precisely because most stochastic phenomena arise through their interaction with nonlinearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.  Meanwhile, the widespread adoption of computational tools and the ability to express theoretical results in numerical algorithms can make any new approaches more readily accessible to application in empirical research. In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuciance and embraced the central role noise plays as the creator of ecological phenomena. Going forward, I hope and predict that we will see greater discussion and use of stochastic phenomena to infer underlying processes in empirical patterns in similar interactions between theoretical and empirical research.

>>>>

What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model,
it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early quantitative ecological models first formulated at the population level
as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
any random component as mere noise [@Coulson2004]. Advances in both theoretical understanding and the advent of widely available and
efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
into quantitative models [@Black2012; @Ovaskainen2010].

@@

The van Kampen system size expansion provides a convenient way to explicitly approximate the "macroscopic" (i.e. population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].  In addition to illustrating how common stochastic models (in particular, Gaussian noise) arise from lower-level descriptions, this result will also provide us a useful framework for interpreting the behavior of stochastic models more generally.  In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:

@@

Note that like the equation for the mean, we have a dynamic equation for the time-evolution of the variance in Eq \eqref{bd}.  We can solve these equations at steady-state by setting both equations zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally to the system size, $N$, which explains the standard deviation increases only as $\sqrt{N}$ while the steady state itself increases in direct proportion to $N$, resulting in the smaller relative size of fluctuations in the larger system in Figure 1.  We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.  For instance, in a logistic model with $b = rn$ and $d = rn^2/N$, we instead find the variance at steady state is $\sigma^2 = K$; significantly larger than in the Levins model.

@@

\begin{equation}
\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d \label{env_noise}
\end{equation}

where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:

@@

Note that the contribution of environmental stochasticity is independent of system size and directly proportional to the variance in the environment.  This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$. This environmental variation no doubt also arise from some more detailed lower level process, but we need not be explicit about those dynamics -- the expansion tells us that it is sufficient to know the overall variance and the degree of auto-correlation.  Recent work has also highlighted the importance of including both demographic and environmental noise [@Melbourne2008; @Bonsall2004].  The system size expansion still requires the assumption that the noise is small relative to the system size.  In the case of large fluctuations, this approximation must be replaced with a different approach, such as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of $\tau_c$, or the degree of auto-correlation, which emerges from our derivation above, has also been a subject of significant interest in the study of stochastic population dynamics [e.g.@Rougharden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].   Early work rarely often assumed environmental noise would be uncorrelated as a mathematical convenience.  As the derivation of Eq \eqref{env_noise} makes clear, for a continuous-time process, the relevant metric is the timescale of auto-correlation in the environmental noise process relative to the timescale (auto-correlation) of the population dynamics.  The Gaussian noise that emerges out of the system-size approximation is always auto-correlated to some degree -- if we measure our environment at consecutive times infinitesimally far apart, we expect nearly identical values.  At intervals much longer than the correlation time, consecutive measurements are effectively independent: auto-correlated noise is all about _relative_ timescales.

## Other forms of stochasticity

@@

A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species. Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guaranteed to eventually go extinct under demographic noise.  While extinction is certainly a phenomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself. Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.  In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibrium of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].

@@

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.  This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transitions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.  The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:

@@

where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.  The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.  This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.  The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g. @Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).  This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.  (Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).  Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.  While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intuition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.  While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.

@@

Despite the nearly ubiquitous association of a ball-in-cup model with ecological discussions of resilience and stability, this potential well diagram (Figure 3C) is also a source of confusion on the topic.  As we have seen, the most salient feature of this diagram is the depth of each well: transition probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.  However, it is important to remember that because the potential function is defined by the above integral, it in fact already combines two more elementary and mechanistically precise quantities from the growth equation: the slope of the growth rate (stability) and the distance to the tipping point (size of the basin of attraction).  Both a steeper negative slope or a longer distance to integrate over between stable and unstable points will create a deeper well.

Empirical evidence for stochastic switching in ecological timeseries is limited by the long time scale usually required. Because a single stochastic transition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013]. However, these transitions are most likely to observed in systems with relatively shallow potential wells.  This is mostly likely to occur of some external change is slowly making the potential well shallower and thus transitions more likely -- a phenomenon that has become to be known as "flickering" and suggested as a potential early warning sign of an approaching transition [@Brock2010].  Examples of flickering between alternative stable states have been suggested in data ranging from geological timescales at the end of glaciation in the Pleistocene [@Taylor1993] to ecological scales over a few decades in lake ecosystems [@Wang2012].  Stochastic transitions between bi-stable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonal microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].

@@

Noise can provide novel information in other ways as well.  Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pushing a system away from an equilibrium to reveal its behavior elsewhere in state space. Observing the size and auto-correlation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-dissipation result we found for generic birth death models).  This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifurcation.  This pattern, known as critical slowing down (in reference to the auto-correlation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has received considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].  Increases in two of these early warning indicators, variance and auto-correlation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.  This simulation is based on a mechanistic model developed by @Dai2012 that closely matches the population dynamics they observed in serial dilutions of budding yeast _Saccharomyces cervisiae_ on a sucrose medium. Because the relevant proteins are located in the extra-cellular matrix rather than inside the cells, budding yeast break down sucrose collaboratively, creating an Allee effect below which the population is too small to sustain itself.  Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing auto-correlation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.

As we saw in the equations from the system-size approximation, the variance and auto-correlation in question are technically properties of an ensemble, and should be computed over replicate simulations.  While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.  If the environmental change is sufficiently slow relative to the measurements, one can assume that the system is close to its stationary state over a given window in time, computing these averages over a rolling time window as we illustrate here, rather than across an ensemble of replicates.

@@

The past two decades have seen widespread recognition of the central role that stochasticity plays through the use of mechanistic models can have in creating and understanding ecological phenomena [@Coulson2004; @Black2012]. We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convenient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].  These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought deterministic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.  Yet too many of these advances have remained limited to the domain of theory; limited to deductive models which may seek to mimic observed patterns but are rarely used to confront empirical data directly.  There is good reason for this: statistical inference of even relatively simple models with multiple sources of noise will always be difficult given limited ecological data [@Knape2011].  However, many stochastic phenomena may suggest new ways to infer rather than merely obscure underlying processes.

The eruption of interest in early warning signals for critical transitions [@Scheffer2009] provides an excellent illustration of this potential.  A range of stochastic phenomena have been suggested and a tested for their potential to reveal subtle signs of ecosystem change prior to a dramatic shift, and prompted a rich dialog between further theory and new experiment. Changes in stochastic properties such as variance or auto-correlation provide evidence of resilience loss that would be invisible to a deterministic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transition.  In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of empirical investigation and further theoretical work to test the potential of this approach.  The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transition or fold bifurcation.

Precisely because most stochastic phenomena arise through their interaction with non-linearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.  Meanwhile, the widespread adoption of computational tools and the ability to express theoretical results in numerical algorithms can make any new approaches more readily accessible to application in empirical research. In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuisance and embraced the central role noise plays as the creator of ecological phenomena. Going forward, I hope and predict that we will see greater discussion and use of stochastic phenomena to infer underlying processes in empirical patterns in similar interactions between theoretical and empirical research.
